---
title: "Characterizing Reactions using MINVAL"
author: "Daniel Osorio, Janneth Gonzalez and Andres Pinzon"
#date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Characterizing Reactions using MINVAL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, dev = "png", fig.width = 7, fig.height = 5, message = FALSE, warning = FALSE)
```
### Characterize Reactions
Characterize the stoichiometric reactions of a metabolic model is a required and time-consuming work. The **MINVAL** package includes the `characterizeReactions` function to characterize the stoichiometric functions and metabolites by compartment.

To load the **MINVAL** package just type:
```{r}
library(minval)
```
To show the potential use of the `characterizeReactions` function of the **MINVAL** package the Human Metabolic Reconstruction (RECON 2.04) was included in a human-readable format. To load it just type:
```{r}
RECON <- read.csv(system.file("extdata", "rRECON2.csv", 
                              package = "minval"))
dim(RECON)
```
The `characterizeReactions` function requires a set of stoichiometric reactions as input. The given stoichiometric reactions must have the following mandatory characteristics:

- Arrows symbols must be given in the form `'=>'` or `'<=>'` (Inverse arrow symbols `'<='` or other types as: `'-->'`, `'<==>'`, `'->'` will not be parsed and will lead to errors.)
- Arrow symbols and plus signs (`+`) must be surrounded by a space character.
- Stoichiometric coefficients must be surrounded by a space character and not by parentheses.
- Each metabolite must have only one stoichiometric coefficient, substituents must be joined to metabolite name by a hyphen (`-`) symbol.
- Exchange reactions must have only one metabolite before arrow symbol.
- Compartments must be given between square brackets (`[compartment]`) joined at the end of metabolite name.

The `characterizeReactions` function: 
- Counts the number of reactions
- Computes the relative frequency of each reaction type (transport, exchange and compartmentalized)
- Computes the relative frequency of reactions by compartment
- Counts the number of unique metabolites
- Computes the relative frequency of metabolites by compartment

Finally the `characterizeReactions` function returns all these information as a labeled list. To characterize a set of stoichiometric reactions just type:
```{r}
RECONcharacteristics <- characterizeReactions(reactionList = RECON$REACTION)
RECONcharacteristics
```
Computed values can be easy plotted as follows:
```{r, echo = FALSE}
compartmentNames <- c(c = "Cytosol", 
                      e = "Extracellular", 
                      g = "Golgi apparatus", 
                      l = "Lysosome", 
                      m = "Mitochondria", 
                      n = "Nucleus", 
                      r = "Endoplasmic reticulum",
                      x = "Peroxisome"
                      )
```
#### Type of Reactions
Metabolic models generally includes three types of reactions:

- **Compartmentalized reactions:** Reactions in where all involved metabolites (all reactants and products) were assigned to the same compartment. e.g. `h[m] + nadph[m] + o2[m] + 25hvitd2[m]  => h2o[m] + nadp[m] + 1a25dhvitd2[m]`
- **Transport reactions:** Reactions in where the involved metabolites (reactants or products) were assigned to two or more compartments. e.g. `2 hco3[e] + na1[e] <=> 2 hco3[c] + na1[c]`
- **Exchange reactions:** Reaction used to import or release metabolites to the boundary. e.g. `acetone[e] <=>`

To plot it just type:
```{r}
pie(x = RECONcharacteristics$rType, 
    main = "Reactions by Type"
    )
```

#### Reactions by Compartment
Compartmentalized reactions identified also are categorized and reported as a relative frequency to their associated compartment. To plot it just type:
```{r}
pie(x = RECONcharacteristics$cReaction, 
    main = "Reactions by Compartment", 
    labels = compartmentNames
    )
```

#### Metabolites by Compartment
Metabolites are categorized and reported as a relative frequency to their associated compartment. To plot it just type:
```{r}
pie(x = RECONcharacteristics$cMetabolites, 
    main = "Metabolites by Compartment", 
    labels = compartmentNames
    )
```